Mixture Models for Single-Cell Assays with Application to Vaccine Studies

### 2013-05-23 MIMOSA.tex
More revisions

\section{Results}
\label{s:results}
In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data. We evaluated and compared the performance of MIMOSA against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Newton2004} where a false discovery (for the ICS data) is a day 0 sample (non--responder) that is incorrectly identified as a responder.

\subsection{ICS}
In the ICS data we considered observations from vaccinees and placebos at the day 0 time point and day 182 time point as true negatives, and observations from vaccinees at the day 182 time point as true positives. The MIMOSA model has higher sensitivity and specificity than method compared here for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E.
At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positives being detected.
In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065}, panels B,D,F). Here we present the results based on IL2 and IFN-$\gamma$ alone and the subset IL2 and/or IFN-$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}. These results are consistent for other cytokines and cytokine combinations (see Supplementary Figure 1).  We also evaluated,through simulation, what would happen to the ROC curves when these assumptions about true responders and non-responders did not hold (\textit{i.e.,} when not all day 182 vaccinees are true responders, Supplementary Figure 6), and found that we are potentially underestimating the sensitivity of all methods considered here due to real non-responders at day 182.

@@

We verified the assumption of a common distribution for the $p^{(u)}_{i}$
across subjects by examining the empirical estimates of $p^{(u)}_{i}$ and the
density of the Beta distribution with parameters $\hat{\alpha}^{(u)},
\hat{\beta}^{(u)}$ fitted to ENV-1-PTEG stimulated
CD4+/IFN--$\gamma$+ T--cell ICS data (Supplementary Figure 2).

>>>>

\section{Results}
\label{s:results}
In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data. We evaluated and compared the performance of MIMOSA against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for each data set~\citep{Newton2004} where a false discovery (for the ICS data) is a day 0 sample or placebo recipient (expected non--responder) that is incorrectly identified as a responder.

\subsection{ICS}
In the ICS data we considered observations from vaccinees and placebos at the day 0 time point and day 182 time point as true negatives, and observations from vaccinees at the day 182 time point as true positives. This assumption was required, since we do not have independent, external measurement of the true response for the day 182 vaccinees. Under these assumptions, the MIMOSA model has higher sensitivity and specificity than method compared here for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E.
At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positives being detected.
In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065}, panels B,D,F). Here we present the results based on IL2 and IFN-$\gamma$ alone as well as the subset expressing IL2 or IFN-$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}. These results are consistent for other cytokines and cytokine combinations, including higher order combinations  (see Supplementary Figure 1).  We also evaluated,through simulation, what would happen to the ROC curves when our assumptions about true responders and non-responders did not hold (\textit{i.e.,} when not all day 182 vaccinees are true responders, Supplementary Figure 2), and found that we are potentially underestimating the sensitivity of all methods considered here due to real non-responders at day 182.

@@

We verified the assumption of a common distribution for the $p^{(u)}_{i}$
across subjects by examining the empirical estimates of $p^{(u)}_{i}$ and the
density of the Beta distribution with parameters $\hat{\alpha}^{(u)},
\hat{\beta}^{(u)}$ fitted to ENV-1-PTEG stimulated
CD4+/IFN--$\gamma$+ T--cell ICS data (Supplementary Figure 3). The effect of information sharing on the sensitivity and specificity of the model was also investigated by varying the number of subjects in simulations (Supplementary Figure 4). Our model performed well when as few as 20 subjects were included, although the observed false discovery rate was below the nominal rate with so few subjects.
